<link href="http://kevinburke.bitbucket.org/markdowncss/markdown.css" rel="stylesheet"></link>
A simple example using Ona.R -- good eats
========================================================

ona.R makes is easy to download and work with datasets on [ona](https://ona.io). After downloading, ona.R post-processes your dataset to convert the different columns to the correct type, which it derives from the `type` you specified during the creation of your XLSform. If you haven't read the [basics document](https://github.com/onaio/ona.R/demo/Basics_of_ona.R.html), I recommend that you read that first.

Let us illustrate this with a simple example:

```{r read, warning=FALSE, message=FALSE}
library(ona)
# Download the dataset named good_eats in the account of mberg
good_eats_data <- onaDownload("good_eats", "mberg","mberg")
```

The downloaded data has its fields converted to the right types, something we'll look at in depth later. Before we do that, lets present a quick example of the power of these conversions by looking at dates. Can we look at how much data mberg has been submitting over time?

```{r plot1, warning=FALSE, message=FALSE, fig.height=5, fig.width=7}
# install.packages("ggplot2") if you don't have ggplot2 installed yet
library(ggplot2)
qplot(data=good_eats_data, x=submit_date) + ylab("Submission Count")
```

How about riskiness of the food he is trying and reporting about... has that changed over time?
```{r plot2, warning=FALSE, message=FALSE, fig.height=5, fig.width=7}
qplot(data=good_eats_data, x=submit_date, fill=risk_factor)
```
Looks like it has mberg's getting riskier as time goes on.
